---
title: "Assignment 3 - Differential Expression with edgeR (mobData)"
author: "Md Tariqul Islam"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
    number_sections: false
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4.5, fig.align = "center")
```

# Part A - Setup and data

```{r packages}
# Run once if needed:
# if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install(c("edgeR", "baySeq"))
library(edgeR)
library(baySeq)
```

```{r load-data}
# Relative path: works when this .Rmd sits in the repo root next to /data
load("data/mobData.RData")
head(mobData)
dim(mobData)

mobDataGroups <- c("MM", "MM", "WM", "WM", "WW", "WW")
# MM = triple mutant shoot grafted onto triple mutant root
# WM = wild-type shoot grafted onto triple mutant root
# WW = wild-type shoot grafted onto wild-type root
data(mobAnnotation)
head(mobAnnotation)

d <- DGEList(counts = mobData, group = factor(mobDataGroups))
d
```

# Part B - Lab steps from "Filtering the data"

## Filtering

```{r filter}
dim(d)
d.full <- d                      # keep a copy of the unfiltered object
head(d$counts)
head(cpm(d))
apply(d$counts, 2, sum)          # total counts per sample

# keep loci with > 100 counts-per-million in at least 2 samples
keep <- rowSums(cpm(d) > 100) >= 2
d <- d[keep, ]
dim(d)

# reset library sizes after filtering
d$samples$lib.size <- colSums(d$counts)
d$samples
```

## Normalization (TMM)

```{r normalize}
d <- calcNormFactors(d)
d$samples
```

## Data exploration (MDS)

```{r mds}
plotMDS(d, method = "bcv", col = as.numeric(d$samples$group),
        main = "MDS plot (BCV distance)")
legend("bottomleft", legend = levels(d$samples$group), col = 1:3, pch = 20)
```

## Estimating dispersion - classic (qCML) model

```{r disp-classic}
d1 <- estimateCommonDisp(d, verbose = TRUE)
d1 <- estimateTagwiseDisp(d1)
names(d1)
plotBCV(d1, main = "BCV vs average log CPM (classic model)")
```

## Estimating dispersion - GLM

```{r disp-glm}
design.mat <- model.matrix(~ 0 + d$samples$group)
colnames(design.mat) <- levels(d$samples$group)
design.mat

d2 <- estimateGLMCommonDisp(d, design.mat)
d2 <- estimateGLMTrendedDisp(d2, design.mat, method = "power")
d2 <- estimateGLMTagwiseDisp(d2, design.mat)
plotBCV(d2, main = "BCV vs average log CPM (GLM)")
```

*("Comparing the models in DESeq and edgeR" is skipped, as instructed.)*

## Differential expression - exact test (lab version)

```{r exact-lab}
et12 <- exactTest(d1, pair = c(1, 2))   # WM vs MM
et13 <- exactTest(d1, pair = c(1, 3))   # WW vs MM
et23 <- exactTest(d1, pair = c(2, 3))   # WW vs WM
topTags(et12, n = 10)

de1 <- decideTests(et12, adjust.method = "BH", p.value = 0.05)
summary(de1)
de1tags12 <- rownames(d1)[as.logical(de1)]
plotSmear(et12, de.tags = de1tags12, main = "Smear plot: WM vs MM (exact test)")
abline(h = c(-2, 2), col = "blue")
```

## Differential expression - GLM (lab version)

```{r glm-lab}
fit <- glmFit(d2, design.mat)
lrt12 <- glmLRT(fit, contrast = c(1, -1, 0))
lrt13 <- glmLRT(fit, contrast = c(1, 0, -1))
lrt23 <- glmLRT(fit, contrast = c(0, 1, -1))
topTags(lrt12, n = 10)

de2 <- decideTests(lrt12, adjust.method = "BH", p.value = 0.05)
summary(de2)
de2tags12 <- rownames(d2)[as.logical(de2)]
plotSmear(lrt12, de.tags = de2tags12, main = "Smear plot: MM vs WM (GLM)")
abline(h = c(-2, 2), col = "blue")
```

# Part C - Analyses for the homework questions

All WW vs MM comparisons below are oriented the same way: **logFC = log2(WW / MM)**, so
"downregulated" means lower expression in WW than in MM.

```{r q3-exact-common}
# Q3: exact test, COMMON dispersion, WW vs MM
et_WW_MM <- exactTest(d1, pair = c("MM", "WW"), dispersion = "common")
de_et <- decideTests(et_WW_MM, adjust.method = "BH", p.value = 0.05)
summary(de_et)
topTags(et_WW_MM, n = 10)

n_down_et <- sum(de_et == -1)
n_up_et   <- sum(de_et ==  1)
n_ns_et   <- sum(de_et ==  0)
```

```{r q4-glm}
# Q4: GLM likelihood-ratio test, WW vs MM (contrast WW - MM)
lrt_WW_MM <- glmLRT(fit, contrast = c(-1, 0, 1))
de_glm <- decideTests(lrt_WW_MM, adjust.method = "BH", p.value = 0.05)
summary(de_glm)
topTags(lrt_WW_MM, n = 10)

n_down_glm <- sum(de_glm == -1)
n_up_glm   <- sum(de_glm ==  1)

down_et  <- rownames(d1)[de_et  == -1]
down_glm <- rownames(d2)[de_glm == -1]
up_et    <- rownames(d1)[de_et  ==  1]
up_glm   <- rownames(d2)[de_glm ==  1]
n_down_both <- length(intersect(down_et, down_glm))
n_up_both   <- length(intersect(up_et, up_glm))
n_de_et  <- n_down_et + n_up_et
n_de_both <- n_down_both + n_up_both
pct_shared <- if (n_de_et > 0) round(100 * n_de_both / n_de_et) else NA

comparison <- data.frame(
  Method = c("Exact test (common dispersion)", "GLM LRT (tagwise dispersion)", "Shared by both"),
  Down   = c(n_down_et, n_down_glm, n_down_both),
  Up     = c(n_up_et,   n_up_glm,   n_up_both))
knitr::kable(comparison, caption = "WW vs MM, BH-adjusted p < 0.05")

# Note: the lab's lrt13 uses contrast c(1,0,-1) = MM - WW, so its "Down" and "Up"
# rows are swapped relative to this table. The genes are the same.
```

```{r q2-numbers, include=FALSE}
common_bcv <- sqrt(d1$common.dispersion)
tag_bcv    <- sqrt(d1$tagwise.dispersion)
rho        <- cor(d1$AveLogCPM, tag_bcv, method = "spearman")
trend_word <- ifelse(rho < 0, "decreases", "increases")
low  <- d1$AveLogCPM <= quantile(d1$AveLogCPM, 0.25)
high <- d1$AveLogCPM >= quantile(d1$AveLogCPM, 0.75)
med_bcv_low  <- median(tag_bcv[low])
med_bcv_high <- median(tag_bcv[high])
```

```{r q5-ql}
# Q5: a different edgeR method - quasi-likelihood F-test (glmQLFit / glmQLFTest)
qlfit <- glmQLFit(d2, design.mat)
qlf_WW_MM <- glmQLFTest(qlfit, contrast = c(-1, 0, 1))
de_ql <- decideTests(qlf_WW_MM, adjust.method = "BH", p.value = 0.05)
summary(de_ql)
n_down_ql <- sum(de_ql == -1)
n_up_ql   <- sum(de_ql ==  1)
```

# Homework answers

## Question 1 - What are SL239 and SL240?

```{r q1}
data.frame(Sample = colnames(d), Group = d$samples$group,
           lib.size = d$samples$lib.size)
```

SL239 and SL240 are the two biological replicate libraries of the **WW** group:
small-RNA sequencing samples from *Arabidopsis* plants in which a **wild-type shoot was
grafted onto a wild-type root**. They are the control (fully wild-type) condition of the
grafting experiment, against which the MM samples (SL236, SL260; triple-mutant shoot on
triple-mutant root) and the WM samples (SL237, SL238; wild-type shoot on triple-mutant
root) are compared. Each column holds the counts for the 3,000 small-RNA loci in that sample.

## Question 2 - Describe the CPM vs BCV plot

```{r q2-plot, echo=FALSE}
plotBCV(d1, main = "BCV vs average log CPM")
```

The plot shows how biological variability relates to abundance for the `r nrow(d1)` loci
kept after filtering. The x-axis is average expression (log2 counts per million) and the
y-axis is the biological coefficient of variation (BCV), the square root of the
negative-binomial dispersion. BCV is the relative variation between biological replicates.
The red line is the common BCV of `r round(common_bcv, 3)`, meaning that a typical locus
varies by about `r round(100 * common_bcv)`% between replicates. The black points are the
tagwise (gene-specific) BCVs.

BCV `r trend_word` as abundance rises (Spearman rho = `r round(rho, 2)`; median BCV
`r round(med_bcv_low, 3)` in the lowest-abundance quartile vs `r round(med_bcv_high, 3)` in
the highest). The spread of points is widest at low CPM, where small counts make
estimates noisy. Some loci lie well above the common line, so variability differs from
locus to locus. This supports filtering out low-count loci and using tagwise (or trended)
dispersion rather than one common value for every locus.

*(Check this paragraph against your rendered figure and edit any wording that doesn't match.)*

## Question 3 - Downregulated genes, WW vs MM, exact test with common dispersion

Using `exactTest(d1, pair = c("MM","WW"), dispersion = "common")` and a Benjamini-Hochberg
FDR cutoff of 0.05, **`r n_down_et` loci are downregulated** in WW relative to MM.
Another `r n_up_et` are upregulated, and `r n_ns_et` are not significant.

## Question 4 - Comparison with the GLM

```{r q4-table, echo=FALSE}
knitr::kable(comparison, caption = "WW vs MM, BH-adjusted p < 0.05")
```

The GLM likelihood-ratio test (`glmFit` + `glmLRT`, contrast WW - MM) finds
`r n_down_glm` downregulated and `r n_up_glm` upregulated loci, compared with
`r n_down_et` and `r n_up_et` from the exact test. `r n_de_both` loci
(`r pct_shared`% of the exact-test calls) are significant in the same direction in both.
The counts differ mainly because of how variability is modelled. The exact test applies
one common dispersion to every locus, whereas the GLM uses tagwise dispersions estimated
with the full design matrix. Loci that are more variable than average therefore lose
significance under the GLM, and unusually stable loci gain it. The GLM is also the more
flexible framework: it can handle multiple factors, batch effects and arbitrary contrasts.
The exact test only works for simple two-group comparisons.

*(If you used the lab's `lrt13`, contrast c(1,0,-1) = MM - WW, your "up" and "down"
counts will be swapped. State the direction you used.)*

## Question 5 - New plot with a different edgeR method

For this plot I used edgeR's **quasi-likelihood (QL) F-test** (`glmQLFit` +
`glmQLFTest`), which the edgeR User's Guide recommends for bulk RNA-seq. It is different
from the exact test and the likelihood-ratio test used in the lab. The plot is an
**MD (mean-difference) plot** of the WW vs MM comparison.

```{r q5-plot, echo=FALSE, fig.height=5}
plotMD(qlf_WW_MM, status = de_ql, values = c(1, -1), col = c("red", "blue"),
       legend = "topright",
       main = "MD plot: WW vs MM (quasi-likelihood F-test, FDR < 0.05)")
abline(h = c(-1, 1), col = "darkgrey", lty = 2)
```

Each point is a locus. The x-axis is its average abundance (log2 CPM), and the y-axis is
its log2 fold change (WW / MM). Red points are significantly upregulated in WW
(n = `r n_up_ql`), blue points are significantly downregulated (n = `r n_down_ql`), and
black points are not significant. The dashed lines mark a 2-fold change.

The plot shows that most loci are centered on logFC = 0, which confirms that TMM
normalization worked and that most small-RNA loci do not change between genotypes.
The significant loci are spread across the whole range of abundance, so the
calls aren't driven only by highly expressed loci. At low abundance, larger fold changes
are needed to reach significance because count noise is higher. The QL method also
accounts for uncertainty in the dispersion estimates, so it is usually a little more
conservative than the LRT. That is why it calls `r n_down_ql + n_up_ql` loci here,
compared with `r n_down_glm + n_up_glm` for the GLM LRT. The loci that remain significant
under QL are the most reliable candidates for small RNAs whose abundance depends on the
wild-type versus triple-mutant genotype.

```{r save-q5, include=FALSE}
dir.create("output", showWarnings = FALSE, recursive = TRUE)
png("output/Q5_MDplot_QL_WW_vs_MM.png", width = 1600, height = 1200, res = 200)
plotMD(qlf_WW_MM, status = de_ql, values = c(1, -1), col = c("red", "blue"),
       legend = "topright", main = "MD plot: WW vs MM (QL F-test, FDR < 0.05)")
abline(h = c(-1, 1), col = "darkgrey", lty = 2)
dev.off()
write.csv(topTags(et_WW_MM, n = Inf)$table, "output/WW_vs_MM_exactTest_common.csv")
write.csv(topTags(lrt_WW_MM, n = Inf)$table, "output/WW_vs_MM_glmLRT.csv")
write.csv(topTags(qlf_WW_MM, n = Inf)$table, "output/WW_vs_MM_glmQLF.csv")
```

```{r session}
sessionInfo()
```
