---
title: "Assignment 3 - Differential Expression with edgeR (mobData)"
author: "Md Tariqul Islam"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
params:
  full: false
output:
  pdf_document:
    toc: false
    number_sections: false
  html_document: default
---

```{r setup, include=FALSE}
full <- isTRUE(params$full)   # FALSE = answers-only PDF, TRUE = full workflow
knitr::opts_chunk$set(echo = full, include = full, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4.5, fig.align = "center")
```

`r if (full) "# Part A - Setup and data" else ""`

```{r packages}
# Run once if needed:
# if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install(c("edgeR", "baySeq"))
library(edgeR)
library(baySeq)
```

```{r load-data}
# Relative path: works when this .Rmd sits in the repo root next to /data
load("data/mobData.RData")
head(mobData)
dim(mobData)

mobDataGroups <- c("MM", "MM", "WM", "WM", "WW", "WW")
# MM = triple mutant shoot grafted onto triple mutant root
# WM = wild-type shoot grafted onto triple mutant root
# WW = wild-type shoot grafted onto wild-type root
data(mobAnnotation)
head(mobAnnotation)

d <- DGEList(counts = mobData, group = factor(mobDataGroups))
d
```

`r if (full) '# Part B - Lab steps from "Filtering the data"' else ""`

`r if (full) "## Filtering" else ""`

```{r filter}
dim(d)
d.full <- d                      # keep a copy of the unfiltered object
head(d$counts)
head(cpm(d))
apply(d$counts, 2, sum)          # total counts per sample

# keep loci with > 100 counts-per-million in at least 2 samples
keep <- rowSums(cpm(d) > 100) >= 2
d <- d[keep, ]
dim(d)

# reset library sizes after filtering
d$samples$lib.size <- colSums(d$counts)
d$samples
```

`r if (full) "## Normalization (TMM)" else ""`

```{r normalize}
d <- calcNormFactors(d)
d$samples
```

`r if (full) "## Data exploration (MDS)" else ""`

```{r mds}
plotMDS(d, method = "bcv", col = as.numeric(d$samples$group),
        main = "MDS plot (BCV distance)")
legend("bottomleft", legend = levels(d$samples$group), col = 1:3, pch = 20)
```

`r if (full) "## Estimating dispersion - classic (qCML) model" else ""`

```{r disp-classic}
d1 <- estimateCommonDisp(d, verbose = TRUE)
d1 <- estimateTagwiseDisp(d1)
names(d1)
plotBCV(d1, main = "BCV vs average log CPM (classic model)")
```

`r if (full) "## Estimating dispersion - GLM" else ""`

```{r disp-glm}
design.mat <- model.matrix(~ 0 + d$samples$group)
colnames(design.mat) <- levels(d$samples$group)
design.mat

d2 <- estimateGLMCommonDisp(d, design.mat)
d2 <- estimateGLMTrendedDisp(d2, design.mat, method = "power")
d2 <- estimateGLMTagwiseDisp(d2, design.mat)
plotBCV(d2, main = "BCV vs average log CPM (GLM)")
```

`r if (full) '*("Comparing the models in DESeq and edgeR" is skipped, as instructed.)*' else ""`

`r if (full) "## Differential expression - exact test (lab version)" else ""`

```{r exact-lab}
et12 <- exactTest(d1, pair = c(1, 2))   # WM vs MM
et13 <- exactTest(d1, pair = c(1, 3))   # WW vs MM
et23 <- exactTest(d1, pair = c(2, 3))   # WW vs WM
topTags(et12, n = 10)

de1 <- decideTests(et12, adjust.method = "BH", p.value = 0.05)
summary(de1)
de1tags12 <- rownames(d1)[as.logical(de1)]
plotSmear(et12, de.tags = de1tags12, main = "Smear plot: WM vs MM (exact test)")
abline(h = c(-2, 2), col = "blue")
```

`r if (full) "## Differential expression - GLM (lab version)" else ""`

```{r glm-lab}
fit <- glmFit(d2, design.mat)
lrt12 <- glmLRT(fit, contrast = c(1, -1, 0))
lrt13 <- glmLRT(fit, contrast = c(1, 0, -1))
lrt23 <- glmLRT(fit, contrast = c(0, 1, -1))
topTags(lrt12, n = 10)

de2 <- decideTests(lrt12, adjust.method = "BH", p.value = 0.05)
summary(de2)
de2tags12 <- rownames(d2)[as.logical(de2)]
plotSmear(lrt12, de.tags = de2tags12, main = "Smear plot: MM vs WM (GLM)")
abline(h = c(-2, 2), col = "blue")
```

`r if (full) "# Part C - Analyses for the homework questions" else ""`

`r if (full) 'All WW vs MM comparisons below use **logFC = log2(WW / MM)**, so "downregulated" means lower expression in WW than in MM.' else ""`

```{r q3-exact-common}
# Q3: exact test, COMMON dispersion, WW vs MM
et_WW_MM <- exactTest(d1, pair = c("MM", "WW"), dispersion = "common")
de_et <- decideTests(et_WW_MM, adjust.method = "BH", p.value = 0.05)
summary(de_et)
topTags(et_WW_MM, n = 10)

n_down_et <- sum(de_et == -1)
n_up_et   <- sum(de_et ==  1)
n_ns_et   <- sum(de_et ==  0)
```

```{r q4-glm}
# Q4: GLM likelihood-ratio test, WW vs MM (contrast WW - MM)
lrt_WW_MM <- glmLRT(fit, contrast = c(-1, 0, 1))
de_glm <- decideTests(lrt_WW_MM, adjust.method = "BH", p.value = 0.05)
summary(de_glm)
topTags(lrt_WW_MM, n = 10)

n_down_glm <- sum(de_glm == -1)
n_up_glm   <- sum(de_glm ==  1)

down_et  <- rownames(d1)[de_et  == -1]
down_glm <- rownames(d2)[de_glm == -1]
up_et    <- rownames(d1)[de_et  ==  1]
up_glm   <- rownames(d2)[de_glm ==  1]
n_down_both <- length(intersect(down_et, down_glm))
n_up_both   <- length(intersect(up_et, up_glm))
n_de_et  <- n_down_et + n_up_et
n_de_both <- n_down_both + n_up_both
pct_shared <- if (n_de_et > 0) round(100 * n_de_both / n_de_et) else NA

comparison <- data.frame(
  Method = c("Exact test (common dispersion)", "GLM LRT (tagwise dispersion)", "Shared by both"),
  Down   = c(n_down_et, n_down_glm, n_down_both),
  Up     = c(n_up_et,   n_up_glm,   n_up_both))
knitr::kable(comparison, caption = "WW vs MM, BH-adjusted p < 0.05")

# Note: the lab's lrt13 uses contrast c(1,0,-1) = MM - WW, so its "Down" and "Up"
# rows are swapped relative to this table. The genes are the same.
```

```{r q2-numbers, include=FALSE}
common_bcv <- sqrt(d1$common.dispersion)
tag_bcv    <- sqrt(d1$tagwise.dispersion)
rho        <- cor(d1$AveLogCPM, tag_bcv, method = "spearman")
trend_word <- ifelse(rho < 0, "decreases", "increases")
low  <- d1$AveLogCPM <= quantile(d1$AveLogCPM, 0.25)
high <- d1$AveLogCPM >= quantile(d1$AveLogCPM, 0.75)
med_bcv_low  <- median(tag_bcv[low])
med_bcv_high <- median(tag_bcv[high])
pct_below_glm <- round(100 * mean(d2$tagwise.dispersion < d1$common.dispersion))
```

```{r q5-ql}
# Q5: a different edgeR method - quasi-likelihood F-test (glmQLFit / glmQLFTest)
qlfit <- glmQLFit(d2, design.mat)
qlf_WW_MM <- glmQLFTest(qlfit, contrast = c(-1, 0, 1))
de_ql <- decideTests(qlf_WW_MM, adjust.method = "BH", p.value = 0.05)
summary(de_ql)
n_down_ql <- sum(de_ql == -1)
n_up_ql   <- sum(de_ql ==  1)
ql_status <- c("Down", "NotSig", "Up")[as.vector(de_ql) + 2]
ql_sig    <- qlf_WW_MM$table$logCPM[as.vector(de_ql) != 0]
max_sig_cpm <- round(max(ql_sig), 1)
n_big_up  <- sum(as.vector(de_ql) == 1 & qlf_WW_MM$table$logFC > 5)
```

# Homework answers

## Question 1 - What are SL239 and SL240?

```{r q1, include=TRUE}
data.frame(Sample = colnames(d), Group = d$samples$group,
           lib.size = d$samples$lib.size)
```

SL239 and SL240 are the two biological replicate libraries of the **WW** group:
small-RNA sequencing samples from *Arabidopsis* plants in which a **wild-type shoot was
grafted onto a wild-type root**. They are the control (fully wild-type) condition of the
grafting experiment, against which the MM samples (SL236, SL260; triple-mutant shoot on
triple-mutant root) and the WM samples (SL237, SL238; wild-type shoot on triple-mutant
root) are compared. Each column holds the counts for the 3,000 small-RNA loci in that sample.

## Question 2 - Describe the CPM vs BCV plot

```{r q2-plot, echo=FALSE, include=TRUE}
plotBCV(d1, main = "BCV vs average log CPM")
```

The plot shows how biological variability relates to abundance for the `r nrow(d1)` loci
kept after filtering. The x-axis is average expression (log2 counts per million) and the
y-axis is the biological coefficient of variation (BCV), the square root of the
negative-binomial dispersion. BCV is the relative variation between biological replicates.
The red line is the common BCV of `r round(common_bcv, 3)`, meaning that a typical locus
varies by about `r round(100 * common_bcv)`% between replicates. The black points are the
tagwise (gene-specific) BCVs.

BCV `r trend_word` as abundance rises (Spearman rho = `r round(rho, 2)`; median BCV
`r round(med_bcv_low, 3)` in the lowest-abundance quartile vs `r round(med_bcv_high, 3)` in
the highest). This is the reverse of the pattern often seen in RNA-seq, where low-count genes
are the most variable. The strict filter (CPM > 100 in at least two samples) has already
removed the noisy low-count loci, so the remaining low-abundance loci are quite consistent
between replicates (BCV mostly 0.13-0.25). BCV then rises to about 0.30-0.35 for the most
abundant loci. As a result, the common BCV overstates the variability of most low-abundance
loci and understates it for highly abundant ones. A minority of loci, mostly between log CPM 6
and 11, lie well above the main cloud (BCV up to about 0.65); their replicates disagree
strongly. Overall, variability depends on both abundance and the individual locus, so a single
common dispersion fits these data poorly. A trended or tagwise dispersion is more appropriate,
and the GLM BCV plot above shows a fitted trend that also rises with abundance.

## Question 3 - Downregulated genes, WW vs MM, exact test with common dispersion

Using `exactTest(d1, pair = c("MM","WW"), dispersion = "common")` and a Benjamini-Hochberg
FDR cutoff of 0.05, **`r n_down_et` loci are downregulated** in WW relative to MM.
Another `r n_up_et` are upregulated, and `r n_ns_et` are not significant.

## Question 4 - Comparison with the GLM

```{r q4-table, echo=FALSE, include=TRUE}
knitr::kable(comparison, caption = "WW vs MM, BH-adjusted p < 0.05")
```

The GLM likelihood-ratio test (`glmFit` + `glmLRT`, contrast WW - MM) finds
`r n_down_glm` downregulated and `r n_up_glm` upregulated loci, compared with
`r n_down_et` and `r n_up_et` from the exact test. `r n_de_both` loci
(`r pct_shared`% of the exact-test calls) are significant in the same direction in both.
The counts differ mainly because of how variability is modelled. The exact test applies
one common dispersion to every locus, whereas the GLM uses tagwise dispersions estimated
with the full design matrix. Because significance depends on the
assumed dispersion, individual loci can gain or lose significance between the two methods.
In these data `r pct_below_glm`% of loci have a GLM tagwise dispersion below the common
value`r if (pct_below_glm > 50) paste0(", so for most loci the GLM assumes less noise, which is consistent with it finding more significant loci overall (", n_down_glm + n_up_glm, " vs ", n_de_et, ")") else ""`. The GLM is also the more
flexible framework: it can handle multiple factors, batch effects and arbitrary contrasts.
The exact test only works for simple two-group comparisons.

## Question 5 - New plot with a different edgeR method

For this plot I used edgeR's **quasi-likelihood (QL) F-test** (`glmQLFit` +
`glmQLFTest`), which the edgeR User's Guide recommends for bulk RNA-seq. It is different
from the exact test and the likelihood-ratio test used in the lab. The plot is an
**MD (mean-difference) plot** of the WW vs MM comparison.

```{r q5-plot, echo=FALSE, fig.height=5, include=TRUE}
plotMD(qlf_WW_MM, status = ql_status, values = c("Up", "Down"), col = c("red", "blue"),
       legend = "topright",
       main = "MD plot: WW vs MM (quasi-likelihood F-test, FDR < 0.05)")
abline(h = c(-1, 1), col = "darkgrey", lty = 2)
```

Each point is a locus. The x-axis is its average abundance (log2 CPM), and the y-axis is
its log2 fold change (WW / MM). Red points are significantly upregulated in WW
(n = `r n_up_ql`), blue points are significantly downregulated (n = `r n_down_ql`), and
black points are not significant. The dashed lines mark a 2-fold change.

Most loci sit around logFC = 0. This is consistent with effective TMM normalization and
indicates that most small-RNA loci change little between the two genotypes. The changes are strongly asymmetric:
`r n_up_ql` loci are higher in WW and only `r n_down_ql` are lower. Of the up-regulated loci,
`r n_big_up` have logFC > 5 (more than a 32-fold increase), so they are nearly absent in MM
and abundant in WW. Down-regulated loci change much less, mostly by 2- to 8-fold.
Significant loci occur from the lowest abundance up to about log CPM `r max_sig_cpm`, while
the few most abundant loci are unchanged. The QL method also accounts for uncertainty in the
dispersion estimates, so it is a little more conservative than the LRT: it calls
`r n_down_ql + n_up_ql` loci here, compared with `r n_down_glm + n_up_glm` for the GLM LRT.
Biologically, the many loci that are almost silent in MM but abundant in WW point to a large
set of small RNAs whose production depends on the wild-type genes that are mutated in MM.
The loci that remain significant under QL are the most reliable candidates for further study.

```{r save-q5, include=FALSE}
dir.create("output", showWarnings = FALSE, recursive = TRUE)
png("output/Q5_MDplot_QL_WW_vs_MM.png", width = 1600, height = 1200, res = 200)
plotMD(qlf_WW_MM, status = ql_status, values = c("Up", "Down"), col = c("red", "blue"),
       legend = "topright", main = "MD plot: WW vs MM (QL F-test, FDR < 0.05)")
abline(h = c(-1, 1), col = "darkgrey", lty = 2)
dev.off()
write.csv(topTags(et_WW_MM, n = Inf)$table, "output/WW_vs_MM_exactTest_common.csv")
write.csv(topTags(lrt_WW_MM, n = Inf)$table, "output/WW_vs_MM_glmLRT.csv")
write.csv(topTags(qlf_WW_MM, n = Inf)$table, "output/WW_vs_MM_glmQLF.csv")
```

```{r session}
sessionInfo()
```
